How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Research works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}. Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common. Specifically, the research questions that we are trying to answer are:

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

>>>>

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Previous work involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}. Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common. To address this goal, we answer the following research questions:

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} contains two one sprites, the cat, which is Scratch's default sprite and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}
